Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-02 paper/paper.tex
abstract and acks

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - deal with all HOGG or Anderson or LMA in the text.
% - deal with everything in the google doc.
% - are there any github issues?

@@

@@

%\author{Adrian Price-Whelan}
%\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

\begin{abstract}\noindent % trust me

Conversion of a noisy parallax measurement into a posterior belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the \gaia\ \tgas\ catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information of $1.2$ million \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the \tgas\ catalog entries by a median factor of 1.2, and a 75, 90, and 99 percentile factor of 1.6, 2.3, and 4.9, respectively; and more precise than previous examples of Bayesian distance estimates by a median factor of 1.2, and a 75, 90 and 99 percentile factor of 1.5, 1.6, and 2.1, respectively.
%The precision can be attributed to the statistics concept of shrinkage.
We independently validate our distances by looking at members of a Milky Way star cluster M67, which is not visible as a cluster in the \tgas\ parallax estimates, but appears clearly as a cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for 1.4 million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

We use stars crossmatched in \tgas\ and \tmass\, that also lie within the observing footprint of \panstarrs\ to access the Greene 3D dust model. We require the photometry have real values, and nonzero, real errors, and remove a small selection of \tmass\ stars that have zero color and zero J band magnitude. The full data set is visualized in the \cmd\ in Figure \ref{fig:data}. The left panel shows the point estimates of the color and absolute magnitude, using the point estimate of the parallax from \gaia, and the $1\sigma$ and $2\sigma$ contours of the distribution. The right panel shows a subset of the data with it's associated error bars. The uncertainties in the colors are fairly well behaved, but the large uncertainties in the absolute magnitude are due to the large uncertainties in the parallax measurements.

@@

\BLcomment{Isn't this repeating what was described in the previous paragraph?}
We need to correct the photometry used in our method, both for the generation of the prior and the likelihood, for dust. With the emergence of 3D dust maps, it is now possible to estimate dust corrections to stars within the Milky Way. The challenge is getting a proper estimate of the distance to the star before inferring $\varpi_{\true}$, since we need a dust correction prior to doing our inference. Dust corrections are particularly non trivial for stars with poor parallax signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. Although the dust model is probabilistic, as well as our distance inference, to move forward, we break from a fully Bayesian framework, in which we might infer the distance and the dust simultaneously (more about this in the discussion). Instead, we correct for dust using a point estimate from the 3D dust map \citep{green15} sampled at a distance estimated from iteratively inferring and then sampling our posterior pdf over $\varpi_{\true}$. In more detail, to apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior, we infer more precise parallaxes to all the stars in our sample. We then use this more precise parallax posterior to get a measurement of dust in the 3D dust map for each star. And repeat. This approach minimizes severe distances, and their associated severe dust corrections, which would occur if using the likelihoods alone. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median, for example, to do a light dust correction, because over correcting for dust can shoot the likelihood into an unphysical, different space of the \cmd\ and bias the parallax inference. For each distance, we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (REF TABLE SFD). We then regenerate our prior using the lightly dust corrected photometry. The dust values we obtain in our final iteration, and which we apply to our photometry, are visualized in galactic coordinates in Figure \ref{fig:dust}. After applying a dust correction, we do not update the uncertainties, but keep them the same as before the dust correction, which is wrong and under representing our uncertainties but this leads, again, to a lighter deconvolution. \BLedit{In other words, the features of our inferred \cmd\ are conservatively broad (broader than if we over-estimated the errors which would have lead to excessive deconvolution and unrealistically narrow features).} We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.

@@

\subsection{prior}
Using \xd\ and the MML method described in the methods section, we generate our prior shown in Figure \ref{fig:prior}.
This is the \gaia\ + \tmass\ data set deconvolved with it's noise, so compared with the raw data in Figure \ref{fig:data}, this is a tighter distribution. The left panel shows 1.3M samples from the prior distribution, and the right panel shows the underlying gaussian mixture model these samples are drawn from. The main sequence and red giant branch are much tighter, as well as the red clump. We'll come back to more features in the data below, especially those seen in the posterior distribution.

@@

\subsection{Output Data}
Using a data driven model of the observed \cmd\ we inferred a posterior belief of the true parallax of each star. Given our method, we call our posterior parallax the photometric parallax.
HOGG: Placeholder for the output format for our data products.
RA, DEC, l, b, tgas source ID, tycho-2 source ID, parallax samples, parallax posterior, flag (some fraction of posterior not in window, negative gaia parallax, ...)

@@

In Section~\ref{sec:method} we listed a set of assumptions upon which
the photometric parallax method in this project is based.
Here we return to these assumptions, and discuss them critically in more
detail.
\begin{description}
\item[stationarity] (HOGG needs to clean this paragraph)
  We have assumed that (locally) the morphology of
  the \cmd\ is independent of signal-to-noise; that is, we have assumed that
  both the low and high
  signal-to-noise objects are being drawn from the same distribution.
  This isn't precisely true because nearby stars in \tgas\ are
  mostly main sequence stars, while we can see luminous giants to much larger
  distances. So the red giant branch is a population of
  relatively lower signal to noise objects compared with the main
  sequence. This does not violate the stationarity assumption in
  detail, because the low and high signal to noise objects have to be
  drawn from the same distribution within some feature or patch of the
  space. This is all assuming the Gaia noise model is correct. If the
  uncertainties are over estimated then we could be over deconvolving
  (generating a tighter distribution than the truth) And in some
  extreme over estimate of uncertainties the lower signal to noise
  giants could be pulled to the main sequence. But with the quoted
  uncertainties this seems unlikely. As you move up the giant branch,
  and probe farther giants, the signal to noise will continue to
  diminish, but because the color uncertainties are not large, the S/N
  spread within a patch of the \cmd\ is well mixed. Possibly a
  metallicity trend with distance, placing more nearby (higher S/N)
  objects at slightly different positions on the CMD pulling lower S/N
  objects to that distribution for our prior. But if something has
  small error bars it is good data is good data both for inference and
  generating the prior.

  For very rare kinds of stars. Inference with our prior is bad for
  underrepresented populations in the \cmd. For example, it appears
  there are only $2$ white dwarfs in \gaia\ DR1. With them being very
  underrepresented in the prior, the amplitudes of the gaussians
  representing them is very small, so the WDs are more likely to be
  pulled to mimic a much farther away MS star.
\item[big data] HOGG
\item[noise model] HOGG
\item[dust] Within our probabilistic framework, we improperly handle
  dust. We are taking a point estimate of the probabilistic dust model
  at a quantile of our distance posterior, correcting our photometry
  using this point estimate, and then re-deriving the prior. A full,
  proper account would sample the dust model at distances sampled from
  the posterior, and propagate those samples back into the prior
  during the iterative process. The reason we don't do a full
  probabilistic treatment, is that this proper dust distribution would
  not be a Gaussian distribution, which \xd\ requires. We could
  calculate this more complex dust distribution and estimate it as a
  Gaussian distribution to feed to \xd, however, this is beyond the
  scope of this paper.
\item[mixture of Gaussians] HOGG
\item[no physics] HOGG
\end{description}

@@

\section{Conclusion}
If you want a conclusion, re-read the abstract

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
  Dustin Lang (Toronto) for cross matched 2MASS-TGAS catalogue
  Jo Bovy (Toronto)
  Boris Leistedt (NYU)
  Adrian Price-Whelan (Princeton)
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

>>>>

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - deal with all DWH or LMA in the text.
% - deal with everything in the google doc.
% - are there any github issues?

@@

\author{Adrian Price-Whelan}
\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

\begin{abstract}\noindent % trust me
Conversion of a noisy parallax measurement into a posterior
belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the \gaia\ \tgas\ catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information of $1.2$ million \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the
\tgas\ catalog entries by a median factor of 1.2, and 20 percent (LMA check)
are more precise by a factor of two or more; they are also more precise than
previous examples of Bayesian distance estimates using a spatial prior.
We independently validate our distances by looking at members of a Milky Way star cluster M67, which is not visible as a cluster in the \tgas\ parallax estimates, but appears clearly as a cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for 1.4 million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

We use stars crossmatched in \tgas\ and \tmass\, that also lie within the observing footprint of \panstarrs\ to access the Greene 3D dust model. We require the photometry have real values, and nonzero, real errors, and remove a small selection of \tmass\ stars that have zero color and zero J band magnitude.

@@

\BLcomment{Isn't this repeating what was described in the previous paragraph?}
We need to correct the photometry used in our method, both for the generation of the prior and the likelihood, for dust. With the emergence of 3D dust maps, it is now possible to estimate dust corrections to stars within the Milky Way. The challenge is getting a proper estimate of the distance to the star before inferring $\varpi_{\true}$, since we need a dust correction prior to doing our inference. Dust corrections are particularly non trivial for stars with poor parallax signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. Although the dust model is probabilistic, as well as our distance inference, to move forward, we break from a fully Bayesian framework, in which we might infer the distance and the dust simultaneously (more about this in the discussion). Instead, we correct for dust using a point estimate from the 3D dust map \citep{green15} sampled at a distance estimated from iteratively inferring and then sampling our posterior pdf over $\varpi_{\true}$. In more detail, to apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior, we infer more precise parallaxes to all the stars in our sample. We then use this more precise parallax posterior to get a measurement of dust in the 3D dust map for each star. And repeat. This approach minimizes severe distances, and their associated severe dust corrections, which would occur if using the likelihoods alone. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median, for example, to do a light dust correction, because over correcting for dust can shoot the likelihood into an unphysical, different space of the \cmd\ and bias the parallax inference. For each distance, we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (REF TABLE SFD). We then regenerate our prior using the lightly dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads, again, to a lighter deconvolution. \BLedit{In other words, the features of our inferred \cmd\ are conservatively broad (broader than if we over-estimated the errors which would have lead to excessive deconvolution and unrealistically narrow features).} We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.

@@

@@

\subsection{Output Data}
Using a data driven model of the observed \cmd\ we inferred a posterior belief of the true parallax of each star. Given our method, we call our posterior parallax the photometric parallax.
DWH: Placeholder for the output format for our data products.
RA, DEC, l, b, tgas source ID, tycho-2 source ID, parallax samples, parallax posterior, flag (some fraction of posterior not in window, negative gaia parallax, ...)

@@

In Section~\ref{sec:method} we listed a set of assumptions upon which
the photometric parallax method in this project is based.
Here we return to these assumptions, and discuss them critically in more
detail.
\begin{description}
\item[stationarity] (DWH needs to clean this paragraph)
  We have assumed that (locally) the morphology of
  the \cmd\ is independent of signal-to-noise; that is, we have assumed that
  both the low and high
  signal-to-noise objects are being drawn from the same distribution.
  This isn't precisely true because nearby stars in \tgas\ are
  mostly main sequence stars, while we can see luminous giants to much larger
  distances. So the red giant branch is a population of
  relatively lower signal to noise objects compared with the main
  sequence. This does not violate the stationarity assumption in
  detail, because the low and high signal to noise objects have to be
  drawn from the same distribution within some feature or patch of the
  space. This is all assuming the Gaia noise model is correct. If the
  uncertainties are over estimated then we could be over deconvolving
  (generating a tighter distribution than the truth) And in some
  extreme over estimate of uncertainties the lower signal to noise
  giants could be pulled to the main sequence. But with the quoted
  uncertainties this seems unlikely. As you move up the giant branch,
  and probe farther giants, the signal to noise will continue to
  diminish, but because the color uncertainties are not large, the S/N
  spread within a patch of the \cmd\ is well mixed. Possibly a
  metallicity trend with distance, placing more nearby (higher S/N)
  objects at slightly different positions on the CMD pulling lower S/N
  objects to that distribution for our prior. But if something has
  small error bars it is good data is good data both for inference and
  generating the prior.

  For very rare kinds of stars. Inference with our prior is bad for
  underrepresented populations in the \cmd. For example, it appears
  there are only $2$ white dwarfs in \gaia\ DR1. With them being very
  underrepresented in the prior, the amplitudes of the gaussians
  representing them is very small, so the WDs are more likely to be
  pulled to mimic a much farther away MS star.
\item[big data] DWH
\item[noise model] DWH
\item[dust] Within our probabilistic framework, we improperly handle
  dust. We are taking a point estimate of the probabilistic dust model
  at a quantile of our distance posterior, correcting our photometry
  using this point estimate, and then re-deriving the prior. A full,
  proper account would sample the dust model at distances sampled from
  the posterior, and propagate those samples back into the prior
  during the iterative process. The reason we don't do a full
  probabilistic treatment, is that this proper dust distribution would
  not be a Gaussian distribution, which \xd\ requires. We could
  calculate this more complex dust distribution and estimate it as a
  Gaussian distribution to feed to \xd, however, this is beyond the
  scope of this paper.
\item[mixture of Gaussians] DWH
\item[no physics] DWH
\end{description}

@@

\section{Conclusion}
If you want to read a conclusion, re-read the abstract!

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
  Stephen Feeney (Flatiron),
  Dustin Lang (Toronto),
  Jo Bovy (Toronto)
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.
